
---
title: "GenderBasedWage by Anisha"
author: "Anisha Kaul"
output: html_document
---


# Project Introduction
This project extracts and analyzes data from National Longitudal Survey of Youth, 1997. The data consists of responses of 8984 participants on 67 different factors. 
Through this project, we are trying to find the answer/s to the following question using data analysis:

> Is there a significant difference in income between men and women?  Does the difference vary depending on other factors (e.g., education, profession, criminal history, marriage status, etc.)?  

### Existing Analyses

According to the American Progress: "The '77 cents' formulation is a colloquialism—shorthand for expressing a complex economic truth. Simply put, what it conveys is the fact that, if you average out what all women, working full time, year round, earn and compare that number to what all men working full time, year round, earn, you find that women take home 77 percent of what men do."

This clearly means that there are several factors that affect the wage gap within the two genders can arise. Also, it is possible for the factors to statiscally influence on the gender wage gap and at the same time, not be in any sort of causal relationship with the wage-gap. Through various statistical methods we will filter out those factors that seem to effect the gender gap more that others out of the present 67 given in the data set. 

It is important to notice that its the wage gap *due to gender* we are more interested in, instead of just wage gap amongst individuals.I hypothesize that the following factors affect the wage gap:

- Job type (industry)
  Some occupations are more male dominated than others, which can result in skew the compensation towards the males. For exmaple if we consider an occupation that involves a lot of physical activity, in my opinion, its highly possible that women will be paid far less per unit time due to the general impression of not being able to produce as much output compared to men. 

- No of years into marriage or Caregiving (household.size)
Its possible that during the the first initial years of the marriage, women tend to be occupied with child bearing and household chores until the children are grown enough are independent. During the initial years, the women may discontinue full-time opportunities and instead continue the same job as part-time, earning lesser than before. Men, do not generally have care-giving reponsibilities and can successfully continue earning without disruptions.

- Geographical Location or the Race (race)
  Gender Wage Gap can be influenced by the region or the race of the counter-parts being considered. 

- Gender (gender)
- Spouse's Estimated Salary in the past year (est.spouse.income.last.year)
- Gross Family Income

##Data Summary

Read the survey data and the data-dictionary availiable:
```{r}
survey.data <- read.csv("nlsy97_income/nlsy97_income.csv")

#Data-Dictionary
data.desc <- read.csv("nlsy97_income_variable names.csv")

sample.survey.data <- subset(survey.data, select=colnames(survey.data)[1:3])
head(sample.survey.data, 4)
```

As you can see that the columns of the data need to be named in a way that can ease our data exploration. There is a data-dictionary avaliable to us, stored in the object called data.desc

```{r}
colnames(survey.data) <- data.desc$New.Name
sample.survey.data <- subset(survey.data, select=colnames(survey.data)[1:3])
head(sample.survey.data, 4)

str(survey.data)
```

Since, we cannot analyse all the data, our population data set can consist of only those factors that we held more important after background information gathering regarding gender-wage-gap

- gender
- household.size
- race
- spouse.income.last.year
- industry
- income
- gross.family.income
- college.type

```{r}
multivariate.list <- c("gender", "household.size", "race", "spouse.income.last.year", "industry", "income", "gross.family.income", "college.type")

survey.data <- subset(survey.data, select=multivariate.list)

str(survey.data)
summary(survey.data)
```
The structure of the data shows us some very absurd values such as:
  - A negative value of Household.size
  - Gender labelled as 1/2
  - Race in integers
  ...and many more.
  
It is possible to study the data in its current form, however, pre-processing the data further will be preferred. Caveats of the current strcuture of the data are: 

    * Current form of the data consists of only INT. Appropriate form of data -> Factors
    * Missing values - can cause data to be skewed
    * Top Coded variables
    
#Methadology

### Deriving Non-Noisy, Categorical Data from Integer Data

Below is how we give meaningful data values to the columns that can make our data visualization better.

```{r}
library(plyr)
survey.data <- transform(survey.data, gender = as.factor(mapvalues(gender, c(1,2,0), c("Male","Female", "No Information"))), race = as.factor(mapvalues(race, c(1,2,3,4), c("Black", "Hispanic","Mixed Race (Non-Hispanic)", "Non-Black /Non-Hispanic"))), college.type = as.factor(mapvalues(college.type,c(1,2,3, -3, -4, -5), c("Public", "Private/Non-Profit", "Private/Profit", "Invalid Skip", "Valid Skip", "Non-Interview"))))

unique(survey.data$race)
unique(survey.data$gender)
unique(survey.data$college.type)

str(survey.data)
```

We notice that industry also consists of integer factors, where the industry should actually be a category.

```{r}
#convert integer intervals into Categorical refrences
survey.data$industry[survey.data$industry %in% 170:290] <- "AGRICULTURE, FORESTRY AND FISHERIES"
survey.data$industry[survey.data$industry %in% 370:490] <- "MINING"
survey.data$industry[survey.data$industry %in% 570:690] <- "UTILITIES"
survey.data$industry[survey.data$industry == 770] <- "CONSTRUCTION"
survey.data$industry[survey.data$industry %in% 1070:3990] <- "MANUFACTURING"
survey.data$industry[survey.data$industry %in% 4070:4590] <- "WHOLESALE TRADE"
survey.data$industry[survey.data$industry %in% 4670:5790] <- "RETAIL TRADE"
survey.data$industry[survey.data$industry %in% 5890] <- "ARTS, ENTERTAINMENT AND RECREATIONAL SERVICES"
survey.data$industry[survey.data$industry %in% 6070:6390] <- "TRANSPORTATION AND WAREHOUSING"
survey.data$industry[survey.data$industry %in% 6470:6780] <- "INFORMATION AND COMMUNICATION"
survey.data$industry[survey.data$industry %in% 6870:7190] <- "FINANCE, INSURANCE, AND REAL ESTATE"
survey.data$industry[survey.data$industry %in% 7270:7790] <- "PROFESSIONAL AND RELATED SERVICES"
survey.data$industry[survey.data$industry %in% 7860:8470] <- "EDUCATIONAL, HEALTH, AND SOCIAL SERVICES"
survey.data$industry[survey.data$industry %in% 8560:8690] <- "ENTERTAINMENT, ACCOMODATIONS, AND FOOD SERVICES"
survey.data$industry[survey.data$industry %in% 8770:9290] <- "OTHER SERVICES"
survey.data$industry[survey.data$industry %in% 9370:9590] <- "PUBLIC ADMINISTRATION"
survey.data$industry[survey.data$industry %in% 9670:9890] <- "ACTIVE DUTY MILITARY"
survey.data$industry[survey.data$industry %in% 9950:9990] <- "ACS SPECIAL CODES"
unique(survey.data$industry)

#convert negative integers references to meaningful references from the data-dictionary
survey.data <- transform(survey.data, industry = as.factor(mapvalues(industry, c("-3","-4","-5"), c("INVALID SKIP","VALID SKIP", "NON-INTERVIEW"))))
```

### Deriving Non-Noisy, Categorical Data from Integer Data

For the negative values that we had for the Categorical data, leaving behind a negative integer would not make a big difference would would look absurd while comparing other 'wordly' categories with numbers. However, in the case of continous data, negative number will not only look absurb for variable from income and household size, but will also skew the data. 

Option 1: For the purposes of our analysis, we will convert the missing or the negative numbers from the continuous variables as NA.

Option 2: We can also replace the data with a value that is the median value for the that variable column.




negative.to.na <- c("household.size", "spouse.income.last.year", "gross.family.income", "income")
industry needs to be factored










### Problem 1: Deriving Categorical Data from Integer Data






